---
title: "Fuel Economy and Car Design in R"
author: "Kuangzi Li"
date: 2026-09-22
format:
  html:
    code-fold: false
---

## Question

This post uses R to examine how vehicle weight and transmission type relate to fuel economy in the `mtcars` dataset.

Data: [`mtcars`](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html), bundled with R's `datasets` package. The data were extracted from the 1974 Motor Trend US magazine and are distributed as part of R.

```{r}
#| label: setup-r
#| message: false
#| warning: false
library(ggplot2)
```

## Load and Prepare the Data

The dataset has one row per car model and includes fuel economy, engine, weight, and transmission variables. I turned row names into a regular column and converted the transmission code into readable labels.

```{r}
#| label: load-mtcars
cars <- mtcars
cars$model <- rownames(mtcars)
cars$transmission <- factor(cars$am, levels = c(0, 1), labels = c("Automatic", "Manual"))
cars$cylinders <- factor(cars$cyl)

head(cars[, c("model", "mpg", "wt", "hp", "cylinders", "transmission")])
```

The dataset is small, so a grouped summary is useful before fitting a model.

```{r}
#| label: transmission-summary
aggregate(
  mpg ~ transmission,
  data = cars,
  FUN = function(x) round(c(mean = mean(x), median = median(x), min = min(x), max = max(x)), 2)
)
```

## Visualize Fuel Economy

Heavier cars tend to have lower miles per gallon. Manual cars in this dataset often appear among the lighter, more fuel-efficient cars.

```{r}
#| label: fig-mtcars-weight-mpg
#| fig-cap: "Fuel economy decreases as vehicle weight increases in the mtcars dataset."
ggplot(cars, aes(x = wt, y = mpg, color = transmission, shape = cylinders)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Weight (1000 lbs)",
    y = "Miles per gallon",
    color = "Transmission",
    shape = "Cylinders",
    title = "Fuel economy by weight and transmission"
  ) +
  theme_minimal()
```

## Fit a Linear Model

I fit a simple linear model to estimate fuel economy from weight and transmission type. This is not a causal model, but it helps quantify the pattern in the plot.

```{r}
#| label: fit-model-r
mpg_model <- lm(mpg ~ wt + transmission, data = cars)
summary(mpg_model)
```

The fitted values below compare a light and a heavy car under each transmission type.

```{r}
#| label: model-predictions-r
new_cars <- expand.grid(
  wt = c(2.2, 3.8),
  transmission = levels(cars$transmission)
)
new_cars$predicted_mpg <- round(predict(mpg_model, newdata = new_cars), 1)

new_cars
```

## What I Found

In this dataset, vehicle weight has a strong negative relationship with fuel economy. Manual cars have higher average MPG, but transmission is also mixed with other design features: many manual cars are lighter. After accounting for weight in a simple model, the estimated transmission difference is smaller than the raw grouped average suggests.
